---
title: Euclidean Distance
description: Comparing National energy generation profiles.
title-block-banner: false
date: '2023-11-01'
categories: [R, Vector Analysis, Distance Metrics, Energy Data]
abstract: "Comparing monthly electricity generation vectors^[Data from the [TidyTuesday 06/06/2023 dataset](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-06-06).] across nations to evaluate structural energy-mix alignment relative to Germany."
abstract-title: Summary
toc: true
execute:
  freeze: auto
---

In observational settings where a formal A/B test wasn't set up, vector distance metrics provide a fast heuristic for pre/post analysis-identifying structurally similar peer entities to evaluate whether macro spikes, market shifts, or regulatory changes observed in a primary market mirrored across baseline controls.

$$d(p, q) = \sqrt{\sum_{i=1}^{n} (p_i - q_i)^2}$$

Where $p_i$ and $q_i$ represent electricity generation (TWh) in month $i$ for Germany and the comparison nation, respectively.

### Vectorized Calculation

```{r, sum-squared-differences-across-all-years}
#| code-overflow: wrap
#| warning: false
#| code-fold: true

library(tidyverse)

owid_energy <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-06-06/owid-energy.csv")

countries <- c(
  "CAN", "MEX", "BLZ", "CRI", "SLV", "GTM", "HND",
  "NIC", "PAN", "BHS", "BRB", "CUB", "DOM", "HTI",
  "JAM", "TTO", "ALB", "AND", "ARM", "AUT", "AZE",
  "BLR", "BEL", "BIH", "BGR", "HRV", "CYP", "CZE",
  "DNK", "EST", "FIN", "FRA", "GEO", "USA", "GRC",
  "HUN", "ISL", "IRL", "ITA", "KAZ", "LVA", "LIE",
  "LTU", "LUX", "MLT", "MDA", "MCO", "MNE", "NLD",
  "MKD", "NOR", "POL", "PRT", "ROU", "RUS", "SMR",
  "SRB", "SVK", "SVN", "ESP", "SWE", "CHE", "UKR",
  "GBR", "VAT"
)

electricity_generation <- owid_energy |>
  select(country, iso_code, year, electricity_generation) |>
  filter(year > 2000 & iso_code %in% countries)

target_country <- "DEU" # Germany's 3 letter ISO country code

target_country_generation <- owid_energy |>
  filter(year > 2000 & year < 2021 & iso_code == target_country) |>
  select(year, electricity_generation) |>
  rename(target_generation = electricity_generation)

countries_with_similarity_score <- electricity_generation |>
  left_join(target_country_generation, by = "year") |>
  group_by(country, iso_code) |>
  summarize(euclidean_distance = sqrt(sum((electricity_generation - target_generation)^2, na.rm = TRUE))) |>
  arrange(euclidean_distance)
```

```{r, euclidean-distance}
#| layout-ncol: 2
#| echo: false
#| tbl-cap:
#|   - "**Top 5 Peer Generation Profiles** (Relative to Germany)"
#|   - "**Top 5 Divergent Generation Profiles** (Relative to Germany)"

library(knitr)

colnames(countries_with_similarity_score) <- c("Country", "ISO Code", "$d$")

kable(head(countries_with_similarity_score, n = 5), digits = 2, format.args = list(big.mark = ","), escape = FALSE)
kable(tail(countries_with_similarity_score , n = 5), digits = 2, format.args = list(big.mark = ","), escape = FALSE)
```


```{r, similiar-countries-code}
#| code-overflow: wrap
#| warning: false
#| code-fold: true
library(tidyverse)
library(showtext)
library(htmltools)
library(gghighlight)

showtext_auto()
showtext_opts(dpi = 600)

font_add_google(name = "Roboto", family = "Roboto")
font <- "Roboto"

title <- paste0(
  "<span>Nations Similar to<span style='color:#6929c4;'> Germany</span></span>"
)

subtitle <- paste0(
  "<span>Electric Generation - Terawatt hours (2000-2020)</span>"
)

similar_countries_highlighted_plot <- owid_energy |>
  filter(iso_code %in% countries | iso_code == "DEU") |>
  filter(year > 2000 & year < 2021 & iso_code != "USA") |>
  ggplot(aes(x = year, y = electricity_generation, group = iso_code, color = iso_code)) +
  geom_line() +
  gghighlight(iso_code %in% c("DEU", "FRA", "CAN"),
              use_direct_label = FALSE) +
  labs(
    title = title,
    subtitle = subtitle,
    y = "Terawatt hours (TWh)",
    x = "Year"
  ) +
    scale_color_manual(
    values = c(
      "DEU" = "#6929c4",
      "FRA" = "#1192e8",
      "CAN" = "#198038",
      "Other" = "#D3D3D3"
    ),
    labels = c("DEU" = "Germany", "FRA" = "France", "CAN" = "Canada")
  ) +
  scale_y_continuous(breaks = seq(0, 1100, by = 250)) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_blank(),
    axis.text = element_text(
      family = font,
      size = 8
    ),
    axis.title = element_text(
      family = font,
      size = 8
    ),
    axis.title.x = element_text(
      margin = margin(5, 0, 0, 0, "mm")
    ),
    axis.title.y = element_text(
      angle = 90,
      margin = margin(0, 5, 0, 0, "mm")
    ),
    panel.grid.major = element_line(colour = "#e0e0e0", linewidth = 0.1),
    legend.text = element_text(
      family = font,
      size = 8
    ),
    plot.title = ggtext::element_textbox_simple(
      family = font,
      size = 12,
      margin = margin(10, 0 , 0, 0)
    ),
    plot.subtitle = ggtext::element_textbox_simple(
      family = font,
      size = 10,
      margin = margin(5, 0, 0, 0)
    ),
    plot.margin = margin(5, 5, 5, 5, "mm")
  )
```

```{r, similiar-countries-plot}
#| echo: false
#| fig-dpi: 600
#| out-width: 70%
#| lightbox:
#|   group: plot

similar_countries_highlighted_plot
```

### Technical Considerations

* **Volume vs. Shape Sensitivity:** Unscaled L2 norm penalizes total generation magnitude. To evaluate pure energy *mix composition* independent of economy size, normalize features to Z-scores or percentage shares ($\% \text{ total TWh}$).
* **Alternative Distance Metrics:** Use **Cosine Distance** ($1 - \cos(\theta)$) to isolate directional alignment, or **Dynamic Time Warping (DTW)** for phase-shifted seasonal curves.